---
title: "ConcussionAnalysis"
author: "Ben Lehmann"
date: "2026-08-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 8, fig.height = 5, fig.align = "center")
```

```{r}

library(dplyr)
library(ggplot2)
library(corrplot)
library(reshape2)
library(gridExtra)
library(car)
library(pwr)
library(caret)
library(pROC)
library(rpart)
library(randomForest)
library(glmnet)

theme_set(theme_minimal(base_size = 12))
set.seed(42)
```



# Introduction

This project analyzes a dataset of `r "15,000"` athlete concussion records, each
including an acute SCAT5 (Sport Concussion Assessment Tool, 5th edition) composite
score, days to return-to-play (RTP), sport, and demographic information. The goals
of this project are to:

1. Build an interpretable **low/medium/severe severity scale** from the raw SCAT5
   acute score.
2. Run **exploratory and descriptive statistics** with visualizations.
3. Conduct **hypothesis tests** on group differences and associations.
4. Apply **multiple-testing correction** across many pairwise comparisons.
5. Frame and run a formal **A/B test** contrasting high-contact vs. non-contact sports.
6. Fit and compare **machine learning models** (classification and regression) that
   predict prolonged recovery and days to RTP.


```{r}
df <- read.csv("athlete_concussion_synthetic_15k.csv", stringsAsFactors = FALSE)

df$injury_date <- as.Date(df$injury_date, format = "%m/%d/%Y")
df$return_to_play_date <- as.Date(df$return_to_play_date, format = "%m/%d/%Y")

df$sex <- factor(df$sex)
df$sport <- factor(df$sport)
df$training_center <- factor(df$training_center)
df$prolonged_rtp <- factor(df$prolonged_rtp, levels = c(0, 1), labels = c("No", "Yes"))
df$high_contact_f <- factor(df$high_contact, levels = c(0, 1), labels = c("Non-Contact", "High-Contact"))
df$repeat_concussion_f <- factor(df$repeat_concussion, levels = c(0, 1), labels = c("First", "Repeat"))

str(df, give.attr = FALSE)
```



## Direction check: which way does the SCAT5 acute score run?

Before building any severity labels, we verify empirically whether a *low* or
*high* `scat5_acute` score corresponds to the *worse* clinical outcome, using
`days_to_rtp` and `prolonged_rtp` as ground truth.