---
title: "ConcussionAnalysis"
author: "Ben Lehmann"
date: "2026-08-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 8, fig.height = 5, fig.align = "center")
```


```{r}
library(tidyverse)
library(ggplot2)
library(scales)
library(viridis)
library(corrplot)
library(reshape2)
library(gridExtra)
library(patchwork)
library(car)
library(broom)
library(pROC)
library(caret)
library(randomForest)
library(rpart)
library(nnet)
library(MASS)
library(glmnet)
library(cluster)
library(factoextra)
library(pwr)
library(knitr)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(color = "grey30"),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold")
    )
)

pal <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e",
         "#e6ab02", "#a6761d", "#666666", "#377eb8")

set.seed(42)
```




```{r}
df <- read.csv("athlete_concussion_enriched_30k.csv", stringsAsFactors = FALSE)

df <- df %>%
  mutate(
    injury_date = as.Date(injury_date, format = "%m/%d/%Y"),
    return_to_play_date = as.Date(return_to_play_date, format = "%m/%d/%Y"),
    sex = factor(sex, levels = c("F", "M")),
    team_level = factor(team_level, levels = c("Youth/High School", "Collegiate", "Elite/Pro")),
    sport = factor(sport),
    season = factor(season, levels = c("Spring","Summer","Fall","Winter")),
    injury_setting = factor(injury_setting),
    concussion_mechanism = factor(concussion_mechanism),
    loss_of_consciousness = factor(loss_of_consciousness, levels = c("No","Yes")),
    amnesia_present = factor(amnesia_present, levels = c("No","Yes")),
    imaging_ordered = factor(imaging_ordered, levels = c("No","Yes")),
    prolonged_rtp_f = factor(prolonged_rtp, levels = c(0,1), labels = c("Normal RTP","Prolonged RTP")),
    high_contact_f = factor(high_contact, levels = c(0,1), labels = c("Lower-Contact","High-Contact")),
    pct_of_baseline = round(scat5_acute / baseline_scat5 * 100, 1)
  )

n_athletes <- n_distinct(df$athlete_id)
kable(data.frame(
  Metric = c("Total concussion events", "Unique athletes", "Sports represented",
             "Training centers", "Date range"),
  Value = c(format(nrow(df), big.mark=","), format(n_athletes, big.mark=","),
            n_distinct(df$sport), n_distinct(df$training_center),
            paste(min(df$injury_date, na.rm=TRUE), "to", max(df$injury_date, na.rm=TRUE)))
), caption = "Dataset Overview")
```




# Building the SCAT5 Severity Scale

Rather than use arbitrary cut points on the raw `scat5_acute` score, we anchor severity to **each athlete's own healthy baseline** by computing:

$$\text{Percent of Baseline} = \frac{\text{scat5\_acute}}{\text{baseline\_scat5}} \times 100$$


```{r}
df <- df %>%
  mutate(
    severity_scale_3 = case_when(
      pct_of_baseline < 35 ~ "Severe",
      pct_of_baseline < 55 ~ "Medium",
      TRUE ~ "Low"
    ),
    severity_scale_3 = factor(severity_scale_3, levels = c("Low","Medium","Severe"), ordered = TRUE),

    severity_scale_5 = case_when(
      pct_of_baseline < 25 ~ "Very Severe",
      pct_of_baseline < 35 ~ "Severe",
      pct_of_baseline < 45 ~ "Moderate",
      pct_of_baseline < 55 ~ "Mild",
      TRUE ~ "Minimal"
    ),
    severity_scale_5 = factor(severity_scale_5,
                               levels = c("Minimal","Mild","Moderate","Severe","Very Severe"),
                               ordered = TRUE)
  )

kable(df %>% count(severity_scale_3) %>% mutate(pct = percent(n/sum(n))),
      caption = "3-Level SCAT5 Severity Scale (Low = mildest injury, Severe = worst)")
kable(df %>% count(severity_scale_5) %>% mutate(pct = percent(n/sum(n))),
      caption = "5-Level SCAT5 Severity Scale")
```




```{r}
p1 <- ggplot(df, aes(x = pct_of_baseline, fill = severity_scale_3)) +
  geom_histogram(binwidth = 2, color = "white", alpha = 0.9) +
  geom_vline(xintercept = c(35, 55), linetype = "dashed", color = "grey30") +
  scale_fill_manual(values = c("Low" = "#1b9e77", "Medium" = "#e6ab02", "Severe" = "#d95f02")) +
  labs(title = "Distribution of Acute SCAT5 as % of Individual Baseline",
       subtitle = "Dashed lines mark the Low/Medium/Severe cut points",
       x = "Percent of Baseline SCAT5 Score", y = "Count", fill = "Severity") +
  theme(legend.position = "top")

p2 <- ggplot(df, aes(x = severity_scale_3, y = days_to_rtp, fill = severity_scale_3)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Low" = "#1b9e77", "Medium" = "#e6ab02", "Severe" = "#d95f02")) +
  labs(title = "Days to Return-to-Play by Severity Scale",
       x = "SCAT5 Severity", y = "Days to RTP") +
  theme(legend.position = "none")

p3 <- ggplot(df, aes(x = severity_scale_5, y = scat5_acute, fill = severity_scale_5)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.3) +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "Raw Acute SCAT5 Score by 5-Level Scale",
       x = "Severity (5-level)", y = "Acute SCAT5 Score") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))

p4 <- df %>%
  count(severity_scale_3, prolonged_rtp_f) %>%
  group_by(severity_scale_3) %>%
  mutate(pct = n/sum(n)) %>%
  ggplot(aes(x = severity_scale_3, y = pct, fill = prolonged_rtp_f)) +
  geom_col(position = "stack", alpha = 0.9) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("Normal RTP" = "#377eb8", "Prolonged RTP" = "#d95f02")) +
  labs(title = "Prolonged RTP Rate by Severity Scale", x = "SCAT5 Severity",
       y = "Proportion of Cases", fill = NULL) +
  theme(legend.position = "top")

(p1 + p2) / (p3 + p4)
```



#EDA

```{r}
p1 <- ggplot(df, aes(x = days_to_rtp)) +
  geom_histogram(binwidth = 1, fill = "#377eb8", color = "white", alpha = 0.9) +
  labs(title = "Days to Return-to-Play", x = "Days", y = "Count")

p2 <- ggplot(df, aes(x = scat5_acute)) +
  geom_histogram(binwidth = 2, fill = "#d95f02", color = "white", alpha = 0.9) +
  labs(title = "Acute SCAT5 Score", x = "Score", y = "Count")

p3 <- ggplot(df, aes(x = age_at_injury)) +
  geom_histogram(binwidth = 1, fill = "#1b9e77", color = "white", alpha = 0.9) +
  labs(title = "Age at Injury", x = "Age", y = "Count")

p4 <- ggplot(df, aes(x = bmi)) +
  geom_histogram(binwidth = 0.5, fill = "#7570b3", color = "white", alpha = 0.9) +
  labs(title = "BMI", x = "BMI", y = "Count")

(p1 + p2) / (p3 + p4)
```





```{r}
p1 <- df %>%
  mutate(sport = fct_reorder(sport, days_to_rtp, .fun = median)) %>%
  ggplot(aes(x = sport, y = days_to_rtp, fill = sport)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.25) +
  scale_fill_viridis_d() +
  coord_flip() +
  labs(title = "Days to RTP by Sport", x = NULL, y = "Days to RTP") +
  theme(legend.position = "none")

p2 <- ggplot(df, aes(x = sex, y = scat5_acute, fill = sex)) +
  geom_violin(alpha = 0.8) +
  geom_boxplot(width = 0.15, fill = "white", outlier.alpha = 0.2) +
  scale_fill_manual(values = c("F" = "#e7298a", "M" = "#377eb8")) +
  labs(title = "Acute SCAT5 by Sex", x = "Sex", y = "Acute SCAT5") +
  theme(legend.position = "none")

p3 <- ggplot(df, aes(x = team_level, y = days_to_rtp, fill = team_level)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.25) +
  scale_fill_manual(values = c("Youth/High School" = "#66a61e", "Collegiate" = "#e6ab02", "Elite/Pro" = "#a6761d")) +
  labs(title = "Days to RTP by Competition Level", x = NULL, y = "Days to RTP") +
  theme(legend.position = "none")

p4 <- df %>%
  count(sport, severity_scale_3) %>%
  group_by(sport) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = fct_reorder(sport, pct * (severity_scale_3=="Severe")), y = pct, fill = severity_scale_3)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("Low" = "#1b9e77", "Medium" = "#e6ab02", "Severe" = "#d95f02")) +
  labs(title = "Severity Mix by Sport", x = NULL, y = "Proportion", fill = "Severity") +
  theme(legend.position = "bottom")

(p1 + p2) / (p3 + p4)
```





```{r}
num_vars <- df %>%
  dplyr::select(days_to_rtp, scat5_acute, scat5_pre_rtp, scat5_improvement, baseline_scat5,
                 pct_of_baseline, age_at_injury, total_games_played, total_elite_years,
                 games_per_year, symptom_count_acute, headache_severity_0_10,
                 sleep_hours_post_injury, bmi)

cor_mat <- cor(num_vars, use = "pairwise.complete.obs")
corrplot(cor_mat, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.6, col = colorRampPalette(c("#d95f02","white","#1b9e77"))(200),
         title = "Correlation Matrix: Numeric Concussion Variables", mar = c(0,0,2,0))
```




# Hypothesis Testing

For every test we report the test statistic, degrees of freedom (where relevant), p-value, effect size, and a plain-language interpretation. Significance is judged at $\alpha = 0.05$ unless otherwise corrected.

## Two-Sample Tests: Sex and Contact Level

**H1 — Sex and days to RTP.** $H_0$: mean days-to-RTP is equal for male and female athletes vs. $H_1$: means differ.


```{r}
t_sex <- t.test(days_to_rtp ~ sex, data = df)
w_sex <- wilcox.test(days_to_rtp ~ sex, data = df)

d_sex <- df %>% group_by(sex) %>% summarise(n=n(), mean=mean(days_to_rtp), sd=sd(days_to_rtp))
pooled_sd <- sqrt(((d_sex$n[1]-1)*d_sex$sd[1]^2 + (d_sex$n[2]-1)*d_sex$sd[2]^2) / (sum(d_sex$n)-2))
cohens_d_sex <- (d_sex$mean[2]-d_sex$mean[1]) / pooled_sd

kable(d_sex, digits = 2, caption = "Days to RTP by Sex")
kable(data.frame(
  Test = c("Welch Two-Sample t-test", "Wilcoxon Rank-Sum"),
  Statistic = c(round(t_sex$statistic,3), round(w_sex$statistic,1)),
  df = c(round(t_sex$parameter,1), NA),
  p.value = c(signif(t_sex$p.value,4), signif(w_sex$p.value,4))
), caption = "Sex Difference in Days-to-RTP")
cat("Cohen's d =", round(cohens_d_sex,3))
```



**H2 — Contact level and days to RTP.** $H_0$: mean days-to-RTP equal for high-contact vs. lower-contact sports.

```{r}
t_contact <- t.test(days_to_rtp ~ high_contact_f, data = df)
d_contact <- df %>% group_by(high_contact_f) %>% summarise(n=n(), mean=mean(days_to_rtp), sd=sd(days_to_rtp))
pooled_sd_c <- sqrt(((d_contact$n[1]-1)*d_contact$sd[1]^2 + (d_contact$n[2]-1)*d_contact$sd[2]^2) / (sum(d_contact$n)-2))
cohens_d_contact <- (d_contact$mean[2]-d_contact$mean[1]) / pooled_sd_c

kable(d_contact, digits = 2, caption = "Days to RTP by Contact Level")
kable(tidy(t_contact) %>% dplyr::select(estimate1, estimate2, statistic, parameter, p.value, conf.low, conf.high),
      digits = 4, caption = "Welch t-test: High-Contact vs Lower-Contact")
cat("Cohen's d =", round(cohens_d_contact,3))
```


`r if(t_contact$p.value < 0.05) "**Conclusion:** We reject H0 -- athletes in high-contact sports take significantly longer to return to play." else "**Conclusion:** We fail to reject H0."`


## One-Way ANOVA: Acute SCAT5 Across Sports

$H_0$: mean acute SCAT5 is equal across all 9 sports. $H_1$: at least one sport differs.

```{r}
aov_sport <- aov(scat5_acute ~ sport, data = df)
print(summary(aov_sport))

# Effect size (eta squared)
ss <- summary(aov_sport)[[1]]
eta_sq <- ss[1,"Sum Sq"] / sum(ss[,"Sum Sq"])
cat("\nEta-squared (effect size):", round(eta_sq, 4), "\n")

leveneTest(scat5_acute ~ sport, data = df) %>% tidy() %>% kable(digits=4, caption = "Levene's Test for Homogeneity of Variance")
```



Given the large sample size, the omnibus ANOVA is virtually guaranteed to be significant; the **eta-squared** effect size tells us how *practically* important sport is. We follow up with a Tukey HSD post-hoc test to see which specific sport pairs differ.


```{r}
tukey_res <- TukeyHSD(aov_sport)
tukey_df <- as.data.frame(tukey_res$sport) %>%
  rownames_to_column("comparison") %>%
  arrange(`p adj`)

kable(head(tukey_df, 15), digits = 4, caption = "Top 15 Most Significant Tukey HSD Pairwise Comparisons (Sport)")

tukey_df %>%
  mutate(sig = `p adj` < 0.05, comparison = fct_reorder(comparison, diff)) %>%
  ggplot(aes(x = diff, y = comparison, color = sig)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = lwr, xmax = upr), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c(`TRUE` = "#d95f02", `FALSE` = "grey60")) +
  labs(title = "Tukey HSD 95% CIs: Pairwise Sport Differences in Acute SCAT5",
       x = "Difference in Means", y = NULL, color = "p < 0.05")
```





## Chi-Square Tests of Association

We test whether **prolonged RTP** (binary outcome) is associated with several categorical risk factors.

```{r}
run_chisq <- function(var) {
  tab <- table(df[[var]], df$prolonged_rtp_f)
  test <- chisq.test(tab)
  data.frame(
    Variable = var,
    Chi_sq = round(unname(test$statistic), 2),
    df = unname(test$parameter),
    p.value = signif(test$p.value, 4),
    Cramers_V = round(sqrt(unname(test$statistic) / (sum(tab) * (min(dim(tab)) - 1))), 4)
  )
}

chisq_vars <- c("sex", "high_contact_f", "loss_of_consciousness", "amnesia_present",
                "imaging_ordered", "team_level", "season", "repeat_concussion")
chisq_results <- map_dfr(chisq_vars, run_chisq) %>% arrange(p.value)
kable(chisq_results, caption = "Chi-Square Tests: Association with Prolonged RTP")
```


```{r}
df %>%
  count(loss_of_consciousness, amnesia_present, prolonged_rtp_f) %>%
  group_by(loss_of_consciousness, amnesia_present) %>%
  mutate(pct = n/sum(n)) %>%
  filter(prolonged_rtp_f == "Prolonged RTP") %>%
  ggplot(aes(x = loss_of_consciousness, y = pct, fill = amnesia_present)) +
  geom_col(position = "dodge", alpha = 0.9) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("No" = "#1b9e77", "Yes" = "#d95f02")) +
  labs(title = "Prolonged RTP Rate by Loss of Consciousness and Amnesia",
       x = "Loss of Consciousness", y = "Prolonged RTP Rate", fill = "Amnesia Present")
```

**Interpretation:** Variables with `p.value < 0.05` show a statistically significant association with prolonged recovery. Cramer's V indicates the *strength* of association (values near 0 = weak, near 1 = strong) — this matters because with n = 30,000, even trivial associations can reach significance.



## Correlation Significance Tests

```{r}
cor_test_age <- cor.test(df$age_at_injury, df$scat5_acute, method = "pearson")
cor_test_games <- cor.test(df$total_games_played, df$days_to_rtp, method = "pearson")
cor_test_sleep <- cor.test(df$sleep_hours_post_injury, df$scat5_acute, method = "spearman")

kable(data.frame(
  Pair = c("Age at Injury vs Acute SCAT5", "Career Games Played vs Days to RTP",
           "Post-Injury Sleep vs Acute SCAT5 (Spearman)"),
  r = c(round(cor_test_age$estimate,4), round(cor_test_games$estimate,4), round(cor_test_sleep$estimate,4)),
  p.value = c(signif(cor_test_age$p.value,4), signif(cor_test_games$p.value,4), signif(cor_test_sleep$p.value,4))
), caption = "Correlation Significance Tests")
```


# Multiple Testing Correction

Running many pairwise tests inflates the family-wise Type I error rate. As an example, we test **all pairwise combinations of the 9 sports** on acute SCAT5 score (36 tests) and on days-to-RTP (36 tests) — 72 tests total — and compare uncorrected p-values against **Bonferroni** and **Benjamini-Hochberg (FDR)** corrections.

```{r}
sports <- levels(df$sport)
pairs_grid <- combn(sports, 2, simplify = FALSE)

run_pairwise <- function(pair, outcome) {
  d1 <- df[[outcome]][df$sport == pair[1]]
  d2 <- df[[outcome]][df$sport == pair[2]]
  tt <- t.test(d1, d2)
  data.frame(outcome = outcome, sport1 = pair[1], sport2 = pair[2],
             mean_diff = round(mean(d1) - mean(d2), 3),
             raw_p = tt$p.value)
}

mt_scat5 <- map_dfr(pairs_grid, run_pairwise, outcome = "scat5_acute")
mt_rtp   <- map_dfr(pairs_grid, run_pairwise, outcome = "days_to_rtp")
mt_all <- bind_rows(mt_scat5, mt_rtp) %>%
  mutate(
    bonferroni_p = p.adjust(raw_p, method = "bonferroni"),
    BH_p = p.adjust(raw_p, method = "BH")
  )

summary_tbl <- mt_all %>%
  group_by(outcome) %>%
  summarise(
    n_tests = n(),
    sig_uncorrected = sum(raw_p < 0.05),
    sig_bonferroni = sum(bonferroni_p < 0.05),
    sig_BH_FDR = sum(BH_p < 0.05)
  )
kable(summary_tbl, caption = "Number of Significant Sport-Pair Comparisons Before/After Correction (72 tests total)")

kable(mt_all %>% arrange(raw_p) %>% dplyr::select(outcome, sport1, sport2, mean_diff, raw_p, bonferroni_p, BH_p) %>% head(10),
      digits = 5, caption = "Top 10 Sport-Pair Comparisons by Raw p-value, with Corrections")
```


```{r}
mt_all %>%
  mutate(rank = row_number(), .by = outcome) %>%
  arrange(outcome, raw_p) %>%
  group_by(outcome) %>%
  mutate(rank = row_number()) %>%
  ggplot(aes(x = rank)) +
  geom_point(aes(y = raw_p, color = "Raw p-value"), size = 1.5, alpha = 0.7) +
  geom_point(aes(y = bonferroni_p, color = "Bonferroni"), size = 1.5, alpha = 0.7) +
  geom_point(aes(y = BH_p, color = "Benjamini-Hochberg"), size = 1.5, alpha = 0.7) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey40") +
  facet_wrap(~outcome, scales = "free_x") +
  scale_color_manual(values = c("Raw p-value" = "grey50", "Bonferroni" = "#d95f02", "Benjamini-Hochberg" = "#1b9e77")) +
  labs(title = "P-value Correction Comparison Across 36 Sport-Pair Tests per Outcome",
       x = "Test Rank (sorted by raw p)", y = "P-value", color = NULL) +
  theme(legend.position = "top")
```

**Bonferroni** (which controls family-wise error rate) is the most conservative and shrinks the number of "significant" pairs the most, while **Benjamini-Hochberg** (which controls false discovery rate) retains more true positives while still guarding against inflation from running 72 simultaneous tests.

# A/B Testing: High-Contact vs. Lower-Contact Sports

We formally frame **high-contact sports (Football, Hockey, Lacrosse, Rugby) as "Group B"** and **lower-contact sports (Baseball, Basketball, Flag Football, Soccer, Softball) as "Group A"**, and test whether Group B has a higher **prolonged-RTP rate** — a binary conversion-style outcome, analogous to a product A/B test.

## Experiment Setup
```{r}
ab_tab <- df %>%
  mutate(group = ifelse(high_contact == 1, "B: High-Contact", "A: Lower-Contact")) %>%
  group_by(group) %>%
  summarise(n = n(), prolonged = sum(prolonged_rtp), rate = mean(prolonged_rtp))

kable(ab_tab, digits = 4, caption = "A/B Groups: Sample Sizes and Prolonged-RTP Conversion Rates")
```

## Two-Proportion Z-Test

$H_0: p_A = p_B$ (equal prolonged-RTP rates) vs. $H_1: p_B > p_A$ (one-sided).


```{r}
x <- ab_tab$prolonged[ab_tab$group == "B: High-Contact"]
n_b <- ab_tab$n[ab_tab$group == "B: High-Contact"]
x_a <- ab_tab$prolonged[ab_tab$group == "A: Lower-Contact"]
n_a <- ab_tab$n[ab_tab$group == "A: Lower-Contact"]

prop_test <- prop.test(x = c(x, x_a), n = c(n_b, n_a), alternative = "greater", correct = TRUE)
print(prop_test)

# Effect size: Cohen's h
p1 <- x/n_b; p2 <- x_a/n_a
cohens_h <- 2*asin(sqrt(p1)) - 2*asin(sqrt(p2))
cat("\nCohen's h (effect size):", round(cohens_h, 4), "\n")

# Relative risk & odds ratio
rr <- p1 / p2
odds_ratio <- (p1/(1-p1)) / (p2/(1-p2))
cat("Relative Risk (B vs A):", round(rr, 3), "\n")
cat("Odds Ratio (B vs A):", round(odds_ratio, 3), "\n")
```



## Confidence Interval on the Difference
```{r}
diff_p <- p1 - p2
se_diff <- sqrt(p1*(1-p1)/n_b + p2*(1-p2)/n_a)
ci_low <- diff_p - 1.96*se_diff
ci_high <- diff_p + 1.96*se_diff
cat("Difference in proportions (B - A):", round(diff_p, 4), "\n")
cat("95% CI: [", round(ci_low, 4), ",", round(ci_high, 4), "]\n")
```


## Statistical Power of the Observed Experiment
```{r}
pwr_res <- pwr.2p2n.test(h = cohens_h, n1 = n_b, n2 = n_a, sig.level = 0.05, alternative = "greater")
print(pwr_res)

# Minimum sample size needed to detect a smaller effect (h = 0.05) at 80% power, for planning future experiments
pwr_needed <- pwr.2p.test(h = 0.05, sig.level = 0.05, power = 0.8, alternative = "greater")
cat("\nSample size needed per group to detect a small effect (h=0.05) at 80% power:", ceiling(pwr_needed$n), "\n")
```

```{r}
p1_plot <- ab_tab %>%
  ggplot(aes(x = group, y = rate, fill = group)) +
  geom_col(alpha = 0.9, width = 0.55) +
  geom_errorbar(aes(ymin = rate - 1.96*sqrt(rate*(1-rate)/n), ymax = rate + 1.96*sqrt(rate*(1-rate)/n)), width = 0.15) +
  geom_text(aes(label = percent(rate, accuracy = 0.1)), vjust = -1.8, fontface = "bold") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0,0.15))) +
  scale_fill_manual(values = c("A: Lower-Contact" = "#377eb8", "B: High-Contact" = "#d95f02")) +
  labs(title = "A/B Test: Prolonged-RTP Rate", subtitle = "Error bars = 95% CI",
       x = NULL, y = "Prolonged RTP Rate") +
  theme(legend.position = "none")

p2_plot <- df %>%
  mutate(group = ifelse(high_contact == 1, "B: High-Contact", "A: Lower-Contact")) %>%
  ggplot(aes(x = group, y = days_to_rtp, fill = group)) +
  geom_violin(alpha = 0.8) +
  geom_boxplot(width = 0.12, fill = "white", outlier.alpha = 0.2) +
  scale_fill_manual(values = c("A: Lower-Contact" = "#377eb8", "B: High-Contact" = "#d95f02")) +
  labs(title = "Days to RTP Distribution by Group", x = NULL, y = "Days to RTP") +
  theme(legend.position = "none")

p1_plot + p2_plot
```


**Conclusion:** `r if(prop_test$p.value < 0.05) paste0("The one-sided two-proportion z-test is statistically significant (p = ", signif(prop_test$p.value,4), "), and the 95% CI on the rate difference excludes zero. High-contact sports show a ", round(diff_p*100,1), " percentage-point higher prolonged-RTP rate, an effect the experiment was well-powered to detect (power = ", round(pwr_res$power,3), ").") else "The test did not reach significance at alpha = 0.05."`


